---
title: "methods"
author: "Sierra Klemme"
date: "3/9/2022"
output: html_document
---

```{r setup, include = F}
library(tidyverse)
library(sf)
library(dataRetrieval)
library(nhdplusTools)
library(mapview)
library(devtools)
```

# Methods 

## Site Description 

  For this report, a more in depth analysis of how climate change 
impacts the interactions between ground water
and surface water will be reviewed  
at stations near Mammoth Lakes California along Hot Creek. This
site allows for water movement to be monitored due to the assumptions
that groundwater due to various factors will later re-emerge into terrestrial
water downstream of the groundwater source or aquifer [@kalbus_measuring_2006]. Mammoth Lakes 
has a coincidental set up to show the process of water transfer between
GW and SW due to a grouping of ground water monitors 
upstream of a surface water monitor along Hot Creek. Mammoth
Lakes is at 7,880 feet in elevation with soils of low infiltration
and located in a pre-dominantly Mediterranean climate [web soil & plantmaps]. 
This particular site where factors such as climate, temperature, groundwater
levels, and discharge can be monitored will allow this analysis to compare 
climate changes to groundwater and surface water quantities. 


## Watershed Delineation

```{r}

mapviewOptions(fgb = FALSE)

site_coords_CA1 <- readNWISsite("10265150") %>%
  st_as_sf(coords = c('dec_long_va', 'dec_lat_va'), remove= FALSE, crs = 4326) %>%
  st_transform(crs = 4269)


CAwatershed <- st_read('MammothCA.geojson') %>% dplyr::filter(id== 'globalwatershed') 

mapview(CAwatershed) + mapview(site_coords_CA1)
```

