---
title: "methods"
author: "Sierra Klemme"
date: "3/9/2022"
output: html_document
---

```{r setup, include = F}
library(tidyverse)
library(sf)
library(dataRetrieval)
library(nhdplusTools)
library(mapview)
library(devtools)
```

# Methods 

## Site Description 

  For this report, a more in depth analysis of how climate change 
impacts the interactions between ground water
and surface water will be reviewed  
at stations near Mammoth Lakes California along Hot Creek. This
site allows for water movement to be monitored due to the assumptions
that groundwater due to various factors will later re-emerge into terrestrial
water downstream of the groundwater source or aquifer [@kalbus_measuring_2006].
Mammoth Lakes has a coincidental set up to show the process of water transfer between
GW and SW due to a grouping of ground water monitors 
upstream of a surface water monitor along Hot Creek. Mammoth
Lakes is at 7,880 feet in elevation with soils of low infiltration
and located in a pre-dominantly Mediterranean climate [web soil & plantmaps]. 
This particular site where factors such as climate, temperature, groundwater
levels, and discharge can be monitored will allow this analysis to compare 
climate changes to groundwater and surface water quantities. 

## Analysis Methods

  Since this report was multifaceted there were many types of data that needed
to be collected. Data for groundwater and surface water were pulled from USGS monitors
and the climate data was retrieved from (uh??) The data was filtered to only 
obtain ground water depth in feet, surface water discharge in cubic feet per
second, temperature in Fahrenheit, and precipitation in millimeters starting on 
January 19, 1992 to April 6, 2022. General trends for groundwater 
levels and surface water discharge rates to analyze on their own and then 
by adjusting scales the two variable were compared on the same graph to show
any relations between the two attributes. Once there was a general understanding
of the GW and SW relation the climate variables were introduced. Temperature and 
precipitation were graphed together to show their correlation and allow for other
interpretations to be made. Temperature and precipitation were then compared to
ground water depths to show trends between changing climate variables and the 
variability of the water table. To further analyze how surface discharge is affected 
the baseflow moths of July through October were compared to show ranges of 
discharge and infer water sources for the causes of large or minimal ranges between 
the months. While some trends were evident through the figures it was important
to objectify those relations by using statistical test to quantify the correlations.
Non-parametric tests such as the Mann-Kendall and Sens Slope were used to greater
support or contradict assumptions made from the figures. These tests also denoted
the strength of the relationship between the various variables. These tests were
used to analyze the significance of surface water and ground water changes, the 
correlation between groundwater and surface water, the baseflow ranges, and 
temperature changes throughout the study period. Using 


## Watershed Delineation

```{r, fig.cap= "Figure 1. The watershed delineation of the surface water reference gauge near Mammoth Lakes, California." }

mapviewOptions(fgb = FALSE)

site_coords_CA1 <- readNWISsite("10265150") %>%
  st_as_sf(coords = c('dec_long_va', 'dec_lat_va'), remove= FALSE, crs = 4326) %>%
  st_transform(crs = 4269)


CAwatershed <- st_read('MammothCA.geojson') %>% dplyr::filter(id== 'globalwatershed') 

mapview(CAwatershed) + mapview(site_coords_CA1)
```

The delineated watershed shows the area that Hot Creek receives its water inputs. 
The surface water gauge is Hot Flume near Mammoth Lakes USGS 10265150. The 
ground water monitor being used is upstream of the surface water monitor listed 
as USGS 373930118491602. 
