---
title: "results"
author: "Sierra Klemme"
date: "5/2/2022"
output: html_document
---
# Results  

```{r, include=FALSE}

library(dataRetrieval)
library(tidyverse)
library(trend)
library(here)
library(lubridate)
library(plotly)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)



```

## Data Acquisition 

```{r, include=FALSE}
##Surface water discharge data

site_ca <- "10265150" 
q_ca <- readNWISdv(siteNumbers = site_ca, 
                parameterCd = '00060', 
                startDate = '1992-01-19', 
                endDate = '2022-04-06') %>% 
  rename(q_cfs = X_00060_00003) %>%
  addWaterYear(.) %>%
  select(-agency_cd, -X_00060_00003_cd) %>%
  mutate(Site = "Hot CA Flume SW") %>%
  select(-site_no)

```

```{r, include=FALSE}
##Ground water data

gw_depth <- read_csv('gw_data.csv')

gw_depth <- gw_depth %>%
  mutate(Site = "Hot Creek GW") %>%
  select(-site)

```

```{r, include=FALSE}
##Climate Data

mam_climate <- read_csv('Date_Mammoth_climate.csv')

```

## Individual Variable Trends 

Groundwater Depth 

```{r, fig.cap= "Groundwater depth for water years 1992 through 2022 at a ground water site near Mammoth Lakes."}
ggplot(gw_depth, aes(x = Date, y = water_depth)) + 
  geom_line(color = 'slateblue2') + 
  ylab('Ground Water Depth (feet)') +
  theme_bw()

```

Groundwater depths near Mammoth Lakes, California have been progressively getting
deeper since 1992. The graph shows the greatest trending decline was between 2000 and
2003. There is one trend after 2006 that shows an increase in the water table 
but after that spike in ground water recharge the depth goes down again significantly.
Overall there is a positive relationship between ground water depths and time.

Surface Discharge 

```{r, fig.cap= "Represented above is the daily annual discharge showing the rates for water years between 1992 to 2022."}
ggplot(q_ca, aes(x = Date, y = q_cfs)) + 
  geom_line(color = 'paleturquoise') + 
  ylab('Q (cfs)') +
  theme_bw()

```

Discharge is influenced by many outside sources. The outputs along Hot Creek
range from 30 cubic feet per second (cfs) to a little over 300 cfs. With the 
exception of the occasional spikes around 300 cfs the river shows a decreasing 
trend in annual ranges of discharge. 

## Attribute Analysis 

GW and SW 

```{r, fig.cap= "Here a logistical comparison of groundwater and surface water discharge trends are shown for water years 1992 to 2022."}

sw_gw <- left_join(gw_depth, q_ca, by = "Date")

sw_gw_graph <- ggplot() +
    geom_line(data = sw_gw, aes(x=Date, y=water_depth*3), color = 'slateblue2') +  
    geom_line(data = sw_gw, aes(x=Date, y=q_cfs), color = 'paleturquoise') 

sw_gw_graph + scale_y_continuous(name = "Dicharge (cfs)" , sec.axis = sec_axis(trans=~./3 , name= "Groundwater Depth (feet)")) +
  theme_bw() +
  xlab("Date")

```

Comparing ground water depths and surface water discharge rates can help show 
how surface water can be affected by ground water levels. The double y-axis allows 
the two variable to be compared by annual averages. The two variables are on 
different scales but a small correlation can be seen between increasing water 
table levels and discharge peaks. 

Temperature and Precipitation 

```{r, fig.cap= "Above shows the comparison between temperature and precipitation. The data is the same time frame as the ground water and surfacewater data starting in 1992 to 2022."} 

t_annual <- mam_climate %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(Avrg_T = mean(Avrg_T, na.rm = T)) 

p_annual <- mam_climate %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(Precip = mean(Precip, na.rm = T))

t_p_annual <- cbind(t_annual, p_annual) %>%
  select(-3)

t_p <- ggplot() +
  geom_line(data = t_p_annual, aes(x=year, y= Avrg_T), color = 'lightsalmon3', size= 1.2) +
  geom_line(data = t_p_annual, aes(x=year, y=Precip*10), color = 'darkslategray2', size= 1.2)+
  xlab("Water Year") +
  scale_color_brewer(guide = "legend")

t_p + scale_y_continuous(name = "Temp (F)" , sec.axis = sec_axis(trans=~./10 , name= "Precip (mm)")) +
  theme_bw()

```

This graph compares the trends of temperature and precipitation over the studied 
time frame. The temperature has a consistent pattern with no major spikes in 
either decreasing or increasing temperatures, however it can be noted that there
is a general upward trend in the temperature data. Precipitation is much more
variable and has significant ranges over the years. Looking at the graph some
correlation between the two variables can be seen where there is a dip in temperature
precipitation shows a rise and when there is an increase in temperature there
is a significant decrease in precipitation such as the section of data between 
2010 and 2015. 

GW, Temperature, Precipitation 

```{r, fig.cap= "Ground water levels compared to temperature and precipitation climate variables."}
gw_annual <- gw_depth %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(water_depth = mean(water_depth, na.rm = T)) 

gw_tp_annual <- left_join(gw_annual, t_p_annual)

layout(matrix(c(1,2), 1, 2, byrow = TRUE)) +
plot(gw_tp_annual$water_depth, gw_tp_annual$Avrg_T, col="lightsalmon3", pch = 16, main="Temperature", ylab = "Average Temperature (F)", xlab = "Water Depth (ft)") +
plot(gw_tp_annual$water_depth, gw_tp_annual$Precip, col="darkslategray2", pch = 16, main="Precipitation", xlab = "Water Depth (ft)", ylab = "Precipitation (mm)") 
  
```

Ground water recharge is dependent on many factors but two climate variable that
can influence ground water depth is temperature and precipitation. Comparing the
two variable against the water table level allows for a relationship to be interpreted.
Based on the graphs there is allegedly a positive trend between temperature 
and ground water with some outlying points. Water depth is shown to be increasing
as temperature also increases. Precipitation and ground water do not show any 
distinct trends from the graph so a statistical analysis will be needed
to better understand the relation. The extreme points on the precipitation graph
do show some assumptions of when there is high amounts of precipitation the ground
water depth is less and when there is minimal precipitation the water table depth
tends to be deeper. 


Monthly Water Trends

```{r, include=FALSE}

q_month_means <- q_ca %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year, month) %>%
  summarize(q_mean = mean(q_cfs, na.rm = T))

```


Discharge Trends for Baseflow Months

```{r, fig.cap= "Baseflow is represented by the months between July and October. Here the diffrence between monthly discharge is shown." }

q_base <- q_month_means %>% 
  filter(month %in% c('7','8','9','10'))

ggplot(q_base, aes( x= year, y= q_mean)) +
  geom_point(aes(color = factor(month))) +
  scale_y_log10() +
  ylab('Q (cfs)') +
  xlab('Water Year')+
  scale_color_discrete(name = "Month", labels = c("July", "August", "September", "October"))

```

Baseflow discharge rates are generally supported by ground water sources. 
The graph shows a typical monthly trend where surface water discharges decreases 
in months further away from snow melt and summer rain storms. The latest month
of October never surpasses 70 cfs but July has an extremely wide range of discharge
values over the study period. Overall the earlier months have higher discharge 
values than the later months. Some years there are significant differences between
each of the months such as the data from around 1995 and other years there is 
a minimal change in discharge between the baseflow months such as the data around
2020.

Groundwater Baseflow Months
```{r, fig.cap= "To analyze baseflow trends it is necessary to also view how groundwater levels change after warm temperature months." }
gw_month <- gw_depth %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year, month) %>%
  summarize(water_depth = mean(water_depth, na.rm = T)) 


gw_base <- gw_month %>% 
  filter(month %in% c('7','8','9','10'))

ggplot(gw_base, aes( x= year, y= water_depth)) +
  geom_point(aes(color = factor(month))) +
  scale_y_log10() +
  ylab('Groundwater Depth (feet)') +
  xlab('Water Year')+
  scale_color_discrete(name = "Month", labels = c("July", "August", "September", "October"))

```

To compare groundwater with discharge during baseflow months the averages for 
groundwater for the months of July - October are shown.Throughout the years in this
study it is seen that groundwater during baseflow months has progressively been 
getting deeper.


Groundwater and Q in October 
```{r, fig.cap= "After plotting GW and Q for baseflow months a trend was seen in October. By showing this baseflow month, it allows for a more refined analysis of the relationship."}
gw_10 <- gw_base %>%
  filter(month %in% ('10'))

q_10 <- q_base %>%
  filter(month %in% ('10'))


gw_q <- left_join(gw_10, q_10)

ggplot(gw_q, aes(x= q_mean, y= water_depth, color = year)) +
  geom_point() +
  labs(x= 'Q (cfs)', y= 'Groundwater (ft)') +
  scale_color_continuous() +
  theme_bw()

```
From the baseflow months it seems there were correlations between grounDwater and
discharge particularly in the month of October. On the ground water versus discharge
graph for October it is noticeable that water tables closer to the surface result
in greater discharge values and as groundwater gets deeper the discharge decreases.
In this graph it also shows the trend that over the years it is more common to see
deeper water tables and less discharge in the Hot Creek stream. 

## Statistical Trend Analysis

```{r}
#Surface Water Discharge 

mk.test(log10(q_month_means$q_mean))

sens.slope(q_month_means$q_mean)
```

```{r}
#Ground Water Level

mk.test(log10(gw_depth$water_depth))

sens.slope(gw_depth$water_depth)
```

```{r}
#Baseflow months discharge trends

mk.test(log10(q_base$q_mean))

sens.slope(q_base$q_mean)

```

```{r}
#Baseflow months groundwater depths

mk.test(log10(gw_base$water_depth))

sens.slope(gw_base$water_depth)

```

```{r}
#Temperature

t_test <- gw_tp_annual %>%
  select(-Precip)

test <- t_test %>%
  filter(year <2018)

mk.test(log10(test$Avrg_T))

sens.slope(test$Avrg_T)

```

Using a non-parametric statistical test such as the Mann-Kendall showed which
variables were significantly changing and then using the Sen's Slope test identified
what type of trend the parameter was experiencing. For the analysis parameters 
tested included surface water monthly averages and baseflow (July-October)
monthly averages from 1992 to 2022. Also tested was groundwater depth trends and
temperature for the same time period 1992 to 2022. After running the tests it was 
found that monthly surface water discharge rates were significant with a p-value
of 1.1e-07 and a slope of -0.038. Baseflow months also had a significant 
p-value of 5.3e-05 with a negative trending slope of -0.19. The tests on groundwater
levels were significant but to a minuscule degree. The p-value was less than
2.2e-16 with a positive slope of 0.00028. Temperature was the only parameter to 
give insignificant results with a p-value of 0.23 and a slope of 0.048. 


