---
title: "Final Final"
author: "Sierra Klemme"
site: bookdown::bookdown_site
documentclass: book
bibliography: bib440.bib 
output:
  bookdown::gitbook: default
  #bookdown::pdf_book: default
--- 

# index

Placeholder


## subheader
### sub-sub header 

<!--chapter:end:index.Rmd-->

---
title: "methods"
author: "Sierra Klemme"
date: "3/9/2022"
output: html_document
---

```{r setup, include = F}
library(tidyverse)
library(sf)
library(dataRetrieval)
library(nhdplusTools)
library(mapview)
library(devtools)
```

# Question

How is climate change affecting ground water levels that can influence base flows 
in surface water discharge along Hot Creek near Mammoth Lakes, California?


Describe why you want to answer this question here (1 paragraph is fine)

I chose this research topic because ground water being a source for surface water discharge is a curious process and I think it will be interesting to see if there are some actual affects of on discharge. Incorporating climate change into the question will also show how the trends of ground water recharge are being affected from the lack of precipitation. This type of research could also help future water plans by finding ways to store water underground and transporting it other places to help decrease evaporation and naturally filter the water. 

# Methods 

## Watershed Delineation

*** make look better?

```{r}

site_coords_CA1 <- readNWISsite("10265150") %>%
  st_as_sf(coords = c('dec_long_va', 'dec_lat_va'), remove= FALSE, crs = 4326) %>%
  st_transform(crs = 4269)


CAwatershed <- st_read('Data/MammothCA.geojson') %>% dplyr::filter(id== 'globalwatershed') 

mapview(CAwatershed) + mapview(site_coords_CA1)
```

## Data Acquisition 

```{r}
##Surface water discharge data

site_ca <- "10265150" # USGS site number for Hot C A Flume near Mammoth Lakes
q_ca <- readNWISdv(siteNumbers = site_ca, 
                parameterCd = '00060', 
                startDate = '1983-07-07', 
                endDate = '2021-12-18') %>% 
  rename(q_cfs = X_00060_00003)
```


```{r, eval=FALSE}
##Ground water data---Still need to figure out how to collect data

site_ca_gw <- "373845118574202" # USGS site number for GW gage near Mammoth Lakes
depth_1 <- readNWISuv(siteNumbers = site_ca_gw,
                       parameterCd = '72019',
                startDate = '2021-01-07', 
                endDate = '2021-12-18') %>%
  rename(waterdepth = X_72019_00000)
```

## Data Description

Describe your discharge data in three ways:

1) When did the record of data start at your site?

For discharge the record of data starts on July 10, 1983.

2) When did data record end?

The records end on December 17,2021.

3) Are there gaps in the data? 

The discharge data for the surface water gage is quite complete with consistent daily time steps for all recorded years. The groundwater data however does not have as many data points over the years. This could be due to the fact that groundwater levels do not change dramatically often or if the levels were manually monitored. 

## Data Plots

Surface Water Discharge 

```{r}
ggplot(q_ca, aes(x = Date, y = q_cfs)) + 
  geom_line() + 
  ylab('Q (cfs)')
```

Water Table Plot 

```{r, eval=FALSE}
ggplot(depth_1, aes(x = Date, y = waterdepth)) + 
  geom_line() + 
  ylab('Q (cfs)')
```

### Interpretation 

The discharge plot above shows data between 1983 and 2021. Throughout this time period there are several significant spikes with high discharge amounts around 300 cfs. A decreasing pattern seems to have occurred in the last decade with less spikes. 


<!--chapter:end:1_methods.Rmd-->



`r if (knitr:: is_html_output()) '# Refrences {-}'`


<!--chapter:end:wrapping_up.Rmd-->


# Results  

Placeholder


## Site Analysis  
## Monthly Water Trends
## Trends in September
## Test for Trends
## Trend for March
## All months with minimum and median Q 
## Plot of all months 
## Monthly stat comparison of June

<!--chapter:end:2_results.Rmd-->

