---
title: "Final Final"
author: "Sierra Klemme"
site: bookdown::bookdown_site
documentclass: book
bibliography: bib440.bib 
output:
  bookdown::gitbook: default
  #bookdown::pdf_book: default
--- 

#Introduction 

  Within recent decades groundwater research has increased in significance due to 
its influential interactions with surface water. The 
understanding of the exchanges between groundwater and surface water 
has also gained popularity due to the effects on the global fresh water supply.
Groundwater and surface water (GW, SW) are not independent sources 
but rather influential systems 
that effects several hydrologic characteristics such as baseflow and 
recharge. As any other process, the hydrologic cycle is 
effected by outside factors. The term "hydrogeologic environment" 
coined by hydrogeologist J.Tóth denotes the need to acknowledge the 
effects of topography, geology, and climate to further understand
GW and SW interaction [@sophocleous_interactions_2002]. 
Previous research has often defined base flows of surface water 
systems to be sourced by upstream ground water storage's 
[@brutsaert_long-term_2008]. With this reference, benefits from
looking at global changes in temperature and precipitation include 
ways to predict ground water levels impacts on surface water 
discharge and how the overall water budget is being effected by global 
climate change [@huntington_role_2012].

  As anthropogenic effects become more notable in nature it raises a need 
to understand GW and SW exchanges in order to protect one of our most valuable
resources. With activities such as mining and contaminated runoff it is valuable 
to quantify flows between GW and SW to help mitigate damage to the riparian area
and decrease harm to organisms who may ingest contaminated water 
[@kalbus_measuring_2006]. Further studies of the lag 



##Research Question

How is climate change affecting ground water levels that can influence base flows 
in surface water discharge along Hot Creek near Mammoth Lakes, California? Specifically,
the attributes in consideration are temperature, precipitation, ground water 
depth, and surface water discharge raters. 



<!--chapter:end:index.Rmd-->

---
title: "methods"
author: "Sierra Klemme"
date: "3/9/2022"
output: html_document
---

```{r setup, include = F}
library(tidyverse)
library(sf)
library(dataRetrieval)
library(nhdplusTools)
library(mapview)
library(devtools)
```

# Methods 

## Site Description 

  For this report, a more in depth analysis of how climate change 
impacts the interactions between ground water
and surface water will be reviewed  
at stations near Mammoth Lakes California along Hot Creek. This
site allows for water movement to be monitored due to the assumptions
that groundwater due to various factors will later re-emerge into terrestrial
water downstream of the groundwater source or aquifer [@kalbus_measuring_2006].
Mammoth Lakes has a coincidental set up to show the process of water transfer between
GW and SW due to a grouping of ground water monitors 
upstream of a surface water monitor along Hot Creek. Mammoth
Lakes is at 7,880 feet in elevation with soils of low infiltration
and located in a pre-dominantly Mediterranean climate [web soil & plantmaps]. 
This particular site where factors such as climate, temperature, groundwater
levels, and discharge can be monitored will allow this analysis to compare 
climate changes to groundwater and surface water quantities. 

***also put in what methods used to find results 

## Watershed Delineation

```{r, fig.cap= "Figure 1. The watershed delineation of the surface water reference gauge near Mammoth Lakes, California." }

mapviewOptions(fgb = FALSE)

site_coords_CA1 <- readNWISsite("10265150") %>%
  st_as_sf(coords = c('dec_long_va', 'dec_lat_va'), remove= FALSE, crs = 4326) %>%
  st_transform(crs = 4269)


CAwatershed <- st_read('MammothCA.geojson') %>% dplyr::filter(id== 'globalwatershed') 

mapview(CAwatershed) + mapview(site_coords_CA1)
```

The delineated watershed shows the area that Hot Creek receives its water inputs. 
The surface water gauge is Hot Flume near Mammoth Lakes USGS 10265150. The 
ground water monitor being used is upstream of the surface water monitor listed 
as USGS 373930118491602. 

<!--chapter:end:1_methods.Rmd-->

# Results  

```{r, include=FALSE}

library(dataRetrieval)
library(tidyverse)
library(trend)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)

```

## Data Acquisition 

```{r, include=FALSE}
##Surface water discharge data

site_ca <- "10265150" 
q_ca <- readNWISdv(siteNumbers = site_ca, 
                parameterCd = '00060', 
                startDate = '1992-01-19', 
                endDate = '2022-04-06') %>% 
  rename(q_cfs = X_00060_00003) %>%
  addWaterYear(.) %>%
  select(-agency_cd, -X_00060_00003_cd) %>%
  mutate(Site = "Hot CA Flume SW") %>%
  select(-site_no)

```


```{r, include=FALSE}
##Ground water data

gw_depth <- read_csv('gw_data.csv')

gw_depth <- gw_depth %>%
  mutate(Site = "Hot Creek GW") %>%
  select(-site)

```

```{r, include=FALSE}
##Climate Data

mam_climate <- read_csv('Date_Mammoth_climate.csv')

```

## Attributes

Groundwater Depth 

```{r, fig.width = 6,fig.height=7, fig.cap= "Figure 2. Groundwater depth for water years 1992 through 2022 at a ground water site near Mammoth Lakes."}
ggplot(gw_depth, aes(x = Date, y = water_depth)) + 
  geom_line(color = 'slateblue2') + 
  ylab('Ground Water Depth (feet)') +
  theme_bw()

#how can i make this an interacitve graph
```

Groundwater depths near Mammoth Lakes, California have been progressively getting
deeper since 1992. The graph shows the greatest decline happening between 2000 and
200??. There is one trend after 2005?? that shows an increase in the water table 
but after that spike in ground water recharge the depth goes down again. Overall
there is a positive relationship between ground water depths and time, interpreting 
this positive relationship however it means that the ground water depths are 
getting deeper and are actually loseing water. 

Surface Discharge 

```{r, fig.cap= "Figure 3. Represented above is the daily annual discharge showing the rates for water years between 1992 to 2022."}
ggplot(q_ca, aes(x = Date, y = q_cfs)) + 
  geom_line(color = 'paleturquoise') + 
  ylab('Q (cfs)') +
  theme_bw()
```

Discharge is influenced by many outside sources. According to the discharge values
from the graph above it can be inferred that this is a moderate stream but is on
average a smaller stream. The rates range from 30 cubic feet per second (cfs) to 
a little over 300 cfs. With the exception of these occasional spikes the river
seems to show decreasing annual ranges in discharge. 

## Attribute Analysis 

GW and SW Long

```{r, fig.cap= "Figure 4. Groundwater has proven to affect surface water discharge levels. Here a logistical comparison of groundwater and surface water discharge trends is shown over the water years through 1992 to 2022. The double y-axix allows the two variable to be compared by annual averages."}

sw_gw <- left_join(gw_depth, q_ca, by = "Date")

sw_gw_graph <- ggplot() +
    geom_line(data = sw_gw, aes(x=Date, y=water_depth*3), color = 'slateblue2') +  
    geom_line(data = sw_gw, aes(x=Date, y=q_cfs), color = 'paleturquoise') 

   
sw_gw_graph + scale_y_continuous(sec.axis = sec_axis(~./3)) + 
  theme_bw() + 
  ylab("Discharge (cfs)") +
  xlab("Date")
```

Comparing ground water depths and surface water discharge rates can help show 
how surface water can be affected by ground water levels. During baseflows it is 
known that ground water can be a continuous source when there is less precipitation
events. The two variables are on different scales but a small correlation can be seen
between increasing water table levels and discharge peaks. 

Temperature and Precipitation 

```{r, fig.cap= "Figure 5. Above shows the comparison between temperature (light brown) and precipitation (light blue). The data is the same time frams as the GW and SW starting in 1992 to 2022." }

t_annual <- mam_climate %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(Avrg_T = mean(Avrg_T, na.rm = T)) 

p_annual <- mam_climate %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(Precip = mean(Precip, na.rm = T))

t_p_annual <- cbind(t_annual, p_annual) %>%
  select(-3)

t_p <- ggplot() +
  geom_line(data = t_p_annual, aes(x=year, y= Avrg_T), color = 'lightsalmon3') +
  geom_line(data = t_p_annual, aes(x=year, y=Precip*10), color = 'darkslategray2') +
  xlab("Water Year") +
  scale_color_brewer(guide = "legend")

t_p + scale_y_continuous(name = "Temp (F)" , sec.axis = sec_axis(trans=~./10 , name= "Precip (mm)"))
                                             breaks = c(0,1,2,3,4,5) + 
                                            breaks = c(0,20,40,50) +
  theme_bw()+ 
  ggtitle("Average Annual Temp and Precip")

#how to put in legend
```

Climate attributes also have great affects on ground water levels and surface
water discharge rates. This graph compares the trends of temperature and 
precipitation over the studied time frame.Temperature and precipitation have
correlate to GW and SW because of they're influence on processes such as water 
recharge amounts and evaporation. 

**Look at data why is it a sharp decline at end of graph

GW, Temp, Precip

```{r, fig.cap= "Figure 6. Ground water levels compared to temperature and precipitation climate variables."}
gw_annual <- gw_depth %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(water_depth = mean(water_depth, na.rm = T)) 

gw_tp_annual <- left_join(gw_annual, t_p_annual)

layout(matrix(c(1,2), 1, 2, byrow = TRUE)) +
plot(gw_tp_annual$water_depth, gw_tp_annual$Avrg_T, col="lightsalmon3", pch = 16, main="Temperature", ylab = "Average Temperature (F)", xlab = "Water Depth (ft)") +
plot(gw_tp_annual$water_depth, gw_tp_annual$Precip, col="darkslategray2", pch = 16, main="Precipitation", xlab = "Water Depth (ft)", ylab = "Precipitation (mm)") 
  
```

Ground water recharge is dependent on many factors but two climate variable that
can influence ground water depth is temperature and precipitation. Comparing the
two variable against the water table level allows for a relationship to be interpreted.
Based on the graphs there is allegedly a slightly positive trend between temperature 
and ground water with some outlying points. Precipitation and ground water do not
distinctly show any trends from the graph so a statistical analysis will be needed
to better understand the relation. By comparing ground water to these climate 
characteristics there can be further analysis on what types of trends can be predicted
in times of climate change.


Monthly Water Trends

```{r, include=FALSE}

q_month_means <- q_ca %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year, month) %>%
  summarize(q_mean = mean(q_cfs, na.rm = T))

```


Discharge Trends for baseflow months

```{r, fig.cap= "Figure 7. Baseflow is represented by the months between July and October. Here the diffrence between monthly discharge is shown." }
#**how to edit legend!!
  
q_base <- q_month_means %>% 
  filter(month %in% c('7','8','9','10'))

ggplot(q_base, aes( x= year, y= q_mean)) +
  geom_point(aes(color = factor(month))) +
  scale_y_log10() +
  ylab('Q (cfs)') +
  xlab('Water Year')+
  scale_color_discrete()

```

Baseflow discharge rates are generally supported by ground water sources. The 
differences in potentials between a low flowing stream and groundwater allows
for groundwater to rise back up into the stream. (Find paper that says this better)
The graph shows a typical monthly trend how surface water discharges decreases 
in months further away from snow melt and summer rain storms. The latest month
of October never surpasses 70 cfs but July has an extremely wide range of discharge
values over the study period. 

## Statistical Trend Analysis

```{r}
#Surface Water Discharge 

mk.test(log10(q_month_means$q_mean))

sens.slope(q_month_means$q_mean)
```

```{r}
#Ground Water Level

mk.test(log10(gw_depth$water_depth))

sens.slope(gw_depth$water_depth)
```

```{r}
#Baseflow month trends

mk.test(log10(q_base$q_mean))

sens.slope(q_base$q_mean)
```

```{r}
#Temperature

t_test <- gw_tp_annual %>%
  select(-Precip)

test <- t_test %>%
  filter(year <2018)

mk.test(log10(test$Avrg_T))

sens.slope(test$Avrg_T)

**not signigicant change of increasing temp by 0.05 degrees each year

```

Using a non-parametric statistical test such as the Mann-Kendall showed which
variables were significantly changing and then using the Sen's Slope test identified
what type of trend the parameter was experiencing. For the analysis parameters 
tested included surface water monthly averages and baseflow (July-October)
monthly averages from 1992 to 2022. Also tested was groundwater depth trends and
temperature for the same time period 1992 to 2022. After running the tests it was 
found that monthly surface water discharge rates were significant with a p-value
of 1.1e-07 and a slope of -0.038. Baseflow months also had a significant 
p-value of 5.3e-05 with a negative trending slope of -0.19. The tests on groundwater
levels were significant but to a minuscule degree. The p-value was less than
2.2e-16 with a positive slope of 0.00028. Temperature was the only parameter to 
give insignificant results with a p-value of 0.23 and a slope of 0.048. 



<!--chapter:end:2_results.Rmd-->



`r if (knitr:: is_html_output()) '# Refrences {-}'`


<!--chapter:end:wrapping_up.Rmd-->

