---
title: "Final Final"
author: "Sierra Klemme"
site: bookdown::bookdown_site
documentclass: book
bibliography: bib440.bib 
output:
  bookdown::gitbook: default
  #bookdown::pdf_book: default
--- 

#Introduction 

  Within recent decades groundwater research has increased in significance due to 
its influential interactions with surface water. The 
understanding of the exchanges between groundwater and surface water 
has also gained popularity due to the effects on the global fresh water supply.
Groundwater and surface water (GW, SW) are not independent sources 
but rather influential systems 
that effects several hydrologic characteristics such as baseflow and 
recharge. As any other process, the hydrologic cycle is 
effected by outside factors. The term "hydrogeologic environment" 
coined by hydrogeologist J.Tóth denotes the need to acknowledge the 
effects of topography, geology, and climate to further understand
GW and SW interaction [@sophocleous_interactions_2002]. 
Previous research has often defined base flows of surface water 
systems to be sourced by upstream ground water storage's 
[@brutsaert_long-term_2008]. With this reference, benefits from
looking at global changes in temperature and precipitation include 
ways to predict ground water levels impacts on surface water 
discharge and how the overall water budget is being effected by global 
climate change [@huntington_role_2012].

  As anthropogenic effects become more notable in nature it raises a need 
to understand GW and SW exchanges in order to protect one of our most valuable
resources. With activities such as mining and contaminated runoff it is valuable 
to quantify flows between GW and SW to help mitigate damage to the riparian area
and decrease harm to organisms who may ingest contaminated water 
[@kalbus_measuring_2006]. Further studies of the lag 



##Research Question

How is climate change affecting ground water levels that can influence base flows 
in surface water discharge along Hot Creek near Mammoth Lakes, California? Specifically,
the attributes in consideration are temperature, precipitation, ground water 
depth, and surface water discharge raters. 



<!--chapter:end:index.Rmd-->

---
title: "methods"
author: "Sierra Klemme"
date: "3/9/2022"
output: html_document
---

```{r setup, include = F}
library(tidyverse)
library(sf)
library(dataRetrieval)
library(nhdplusTools)
library(mapview)
library(devtools)
```

# Methods 

## Site Description 

  For this report, a more in depth analysis of how climate change 
impacts the interactions between ground water
and surface water will be reviewed  
at stations near Mammoth Lakes California along Hot Creek. This
site allows for water movement to be monitored due to the assumptions
that groundwater due to various factors will later re-emerge into terrestrial
water downstream of the groundwater source or aquifer [@kalbus_measuring_2006].
Mammoth Lakes has a coincidental set up to show the process of water transfer between
GW and SW due to a grouping of ground water monitors 
upstream of a surface water monitor along Hot Creek. Mammoth
Lakes is at 7,880 feet in elevation with soils of low infiltration
and located in a pre-dominantly Mediterranean climate [web soil & plantmaps]. 
This particular site where factors such as climate, temperature, groundwater
levels, and discharge can be monitored will allow this analysis to compare 
climate changes to groundwater and surface water quantities. 

***also put in what methods used to find results 

## Watershed Delineation

```{r}

mapviewOptions(fgb = FALSE)

site_coords_CA1 <- readNWISsite("10265150") %>%
  st_as_sf(coords = c('dec_long_va', 'dec_lat_va'), remove= FALSE, crs = 4326) %>%
  st_transform(crs = 4269)


CAwatershed <- st_read('MammothCA.geojson') %>% dplyr::filter(id== 'globalwatershed') 

mapview(CAwatershed) + mapview(site_coords_CA1)
```


<!--chapter:end:1_methods.Rmd-->

# Results  

```{r, include=FALSE}

library(dataRetrieval)
library(tidyverse)
library(trend)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)

```

## Data Acquisition 

```{r, include=FALSE}
##Surface water discharge data

site_ca <- "10265150" 
q_ca <- readNWISdv(siteNumbers = site_ca, 
                parameterCd = '00060', 
                startDate = '1992-01-19', 
                endDate = '2022-04-06') %>% 
  rename(q_cfs = X_00060_00003) %>%
  addWaterYear(.) %>%
  select(-agency_cd, -X_00060_00003_cd) %>%
  mutate(Site = "Hot CA Flume SW") %>%
  select(-site_no)

```


```{r, include=FALSE}
##Ground water data

gw_depth <- read_csv('gw_data.csv')

gw_depth <- gw_depth %>%
  mutate(Site = "Hot Creek GW") %>%
  select(-site)

```

```{r, include=FALSE}
##Climate Data

mam_climate <- read_csv('Date_Mammoth_climate.csv')

```

## Attributes

Groundwater Depth 

```{r, fig.width = 6,fig.height=7, fig.cap= "Figure 1. Groundwater depth for water years 1992 through 2022 at a ground water site near Mammoth Lakes."}
ggplot(gw_depth, aes(x = Date, y = water_depth)) + 
  geom_line(color = 'slateblue2') + 
  ylab('Ground Water Depth (feet)') +
  theme_bw()
```

Groundwater depths near Mammoth Lakes, California The graph independently represents how the water table has shifted througout the 
years

```{r, fig.cap= "Figure 2. Represented above is the daily annual discharge showing the discharge rates for water years between 1992 to 2022."}
ggplot(q_ca, aes(x = Date, y = q_cfs)) + 
  geom_line(color = 'paleturquoise') + 
  ylab('Q (cfs)') +
  theme_bw()
```


## Attribute Analysis 

GW and SW Long

```{r, fig.cap= "Figure 3. Groundwater has proven to affect surface water discharge levels. Here a logistical comparison of groundwater and surface water discharge trends is shown over the water years through 1992 to 2022. The double y-axix allows the two variable to be compared by annual averages."}

sw_gw <- left_join(gw_depth, q_ca, by = "Date")

sw_gw_graph <- ggplot() +
    geom_line(data = sw_gw, aes(x=Date, y=water_depth*3), color = 'slateblue2') +  
    geom_line(data = sw_gw, aes(x=Date, y=q_cfs), color = 'paleturquoise') 

   
sw_gw_graph + scale_y_continuous(sec.axis = sec_axis(~./3)) + 
  theme_bw() + 
  ylab("Discharge (cfs)") +
  xlab("Date")
```

Temperature and Precipitation 

```{r, fig.cap= "Figure 4. Above shows the comparison between temperature (light brown) and precipitation (light blue). Temperature correlates with precipitation possibilities and how this trend comparison affects evaporation and groundwater affects." }

t_annual <- mam_climate %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(Avrg_T = mean(Avrg_T, na.rm = T)) 

p_annual <- mam_climate %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(Precip = mean(Precip, na.rm = T))

t_p_annual <- cbind(t_annual, p_annual) %>%
  select(-3)

t_p <- ggplot() +
  geom_line(data = t_p_annual, aes(x=year, y= Avrg_T), color = 'lightsalmon3') +
  geom_line(data = t_p_annual, aes(x=year, y=Precip*10), color = 'darkslategray2') +
  xlab("Water Year") +
  scale_color_brewer(guide = "legend")

t_p + scale_y_continuous(name = "Temp (F)" , sec.axis = sec_axis(trans=~./10 , name= "Precip (mm)"))
                                             breaks = c(0,1,2,3,4,5) + 
                                            breaks = c(0,20,40,50) +
  theme_bw()+ 
  ggtitle("Average Annual Temp and Precip")

```

GW, Temp, Precip

```{r, fig.cap= "Figure 5. Temperature and precipictaion have previous evidence that the variables correlate to groundwater level recharge. Therefore the comparison of the dependant variables between temperature and precipitation trends demonstrate how climatic variable affect groundwater levels."}
gw_annual <- gw_depth %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year) %>%
  summarize(water_depth = mean(water_depth, na.rm = T)) 

gw_tp_annual <- left_join(gw_annual, t_p_annual)

layout(matrix(c(1,2), 1, 2, byrow = TRUE)) +
plot(gw_tp_annual$water_depth, gw_tp_annual$Avrg_T, col="lightsalmon3", pch = 16, main="Temperature", ylab = "Average Temperature (F)", xlab = "Water Depth (ft)") +
plot(gw_tp_annual$water_depth, gw_tp_annual$Precip, col="darkslategray2", pch = 16, main="Precipitation", xlab = "Water Depth (ft)", ylab = "Precipitation (mm)") 
  
```

Monthly Water Trends

```{r, include=FALSE}

q_month_means <- q_ca %>%
  dplyr::mutate(month = month(Date),
                year = year(Date)) %>%
  group_by(year, month) %>%
  summarize(q_mean = mean(q_cfs, na.rm = T))

```


Discharge Trends for baseflow months

```{r, fig.cap= "Figure 6. Baseflow is represented by the months between July and October. Here the diffrence between monthly discharge is shown and how surface water discharges decreases in months further away from snwomelt and summer rain storms." }
#**how to edit legend!!
  
q_base <- q_month_means %>% 
  filter(month %in% c('7','8','9','10'))

ggplot(q_base, aes( x= year, y= q_mean)) +
  geom_point(aes(color = factor(month))) + 
  scale_y_log10() +
  ylab('Q (cfs)') +
  xlab('Water Year')+
  scale_color_discrete()

```

## Statistical Trend Analysis

```{r}
#Surface Water Discharge 

mk.test(log10(q_month_means$q_mean))

sens.slope(q_month_means$q_mean)
```

```{r}
#Ground Water Level

mk.test(log10(gw_depth$water_depth))

sens.slope(gw_depth$water_depth)
```

```{r}
#Baseflow month trends

mk.test(log10(q_base$q_mean))

sens.slope(q_base$q_mean)
```

```{r}
#Temperature

t_test <- gw_tp_annual %>%
  select(-Precip)

test <- t_test %>%
  filter(year <2018)

mk.test(log10(test$Avrg_T))

sens.slope(test$Avrg_T)

**not signigicant change of increasing temp by 0.05 degrees each year

```

Using a non-parametric statistical test such as the Mann-Kendall showed which
variables were significantly changing and then using the Sen's Slope test identified
what type of trend the parameter was experiencing. For the analysis parameters 
tested included surface water monthly averages and baseflow (July-October)
monthly averages from 1992 to 2022. Also tested was groundwater depth trends and
temperature for the same time period 1992 to 2022. After running the tests it was 
found that monthly surface water discharge rates were significant with a p-value
of 1.1e-07 and a slope of -0.038. Baseflow months also had a significant 
p-value of 5.3e-05 with a negative trending slope of -0.19. The tests on groundwater
levels were significant but to a minuscule degree. The p-value was less than
2.2e-16 with a positive slope of 0.00028. Temperature was the only parameter to 
give insignificant results with a p-value of 0.23 and a slope of 0.048. 



<!--chapter:end:2_results.Rmd-->



`r if (knitr:: is_html_output()) '# Refrences {-}'`


<!--chapter:end:wrapping_up.Rmd-->

